---
title: "scATAC-seq OpenProblems"
author: "Roger Casals"
date: "2023-11-17"
output: html_document
---

#Data and library reading

```{r}
#############
setwd("~/Github/OpenProblems")

library(arrow)
library(data.table)
library(Matrix)
library(dplyr)
library(tidyr)
library(Seurat)
library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.UCSC.hg38)


train <- read_parquet("multiome_train.parquet")
multi <- read.csv("multiome_var_meta.csv")
meta <- read.csv("multiome_obs_meta.csv")


```


#Create sparse matrix

```{r}
#Create objects with "peaks" and "gene expression"
nueva_train <- train[grep("^chr", train$location), ]
no_chr <- train[!grepl("chr", train$location), ]
train <- nueva_train
```


##Sparse matrix Peaks (normalized counts)

```{r}
customers <- unique(train$location
)
products <- unique(train$obs_id
)

train$row <- match(train$location
                   , customers)
train$col <- match(train$obs_id
                   , products)

df_sparse2 <- sparseMatrix(
  i = train$row, 
  j = train$col,
  x = train$normalized_count,
  dimnames = list(customers,
                  products)
)

dim(df_sparse2)
```


##Sparse matrix Gene expression RNA (normalized_counts)

```{r}

train <- no_chr

customers <- unique(train$location
)
products <- unique(train$obs_id
)

train$row <- match(train$location
                   , customers)
train$col <- match(train$obs_id
                   , products)

df_sparse <- sparseMatrix(
  i = train$row, 
  j = train$col,
  x = train$normalized_count,
  dimnames = list(customers,
                  products)
)

```


I don't know if we should use normalized_counts or counts, because after that in Signac multi-omic processment, there is another normalitzation with IDF (Same that the authors say they do in the preprocessment of the data)


#Follow Signac Multi-omics tutorial

Fragments file has already been created, it is the file with the "intervals" in BED format, the cell id and the "counts".


## RNA OBJECT

```{r}

####### RNA OBJECT
pbmc <- CreateSeuratObject(
  counts = df_sparse,
  assay = "RNA",
  meta.data=meta
)


```


## ATAC OBJECT

```{r}

fragpath <- "~/Github/OpenProblems/nuevo_archivo.tsv.gz"

annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevels(annotation) <- paste0('chr', seqlevels(annotation))

###### PEAKS OBJECT

# create ATAC assay and add it to the object
pbmc[["ATAC"]] <- CreateChromatinAssay(
  counts = df_sparse2,
  sep = c(":", "-"),
  fragments = fragpath,
  annotation = annotation
)
```


### QC


```{r}
DefaultAssay(pbmc) <- "ATAC"

#Nucleosome signal doesn't work, because the widths are bigger than the nucleosome widths that the function has established.
pbmc <- NucleosomeSignal(pbmc)
#See the nucleosomeSignal output, only 1 and Inf.


#Compute the TSSEnrichment
pbmc <- TSSEnrichment(pbmc)

pbmc[['RNA']] <- split(pbmc[["RNA"]], f = pbmc$cell_type)

#See the different 
VlnPlot(
  object = pbmc,
  features = c("nCount_RNA", "nCount_ATAC", "TSS.enrichment"),
  ncol = 4,
  pt.size = 0
)


DensityScatter(pbmc, x = 'nCount_ATAC', y = 'TSS.enrichment', log_x = TRUE, quantiles = TRUE)

summary(pbmc$TSS.enrichment)


pbmc[['RNA']] <- JoinLayers(pbmc[["RNA"]])
DefaultAssay(pbmc) <- "RNA"

pbmc <- JoinLayers(pbmc)






```



#### Filtering cells


```{r}
########################
# filter out low quality cells, changeable numbers, 
pbmc <- subset(
  x = pbmc,
  subset = nCount_ATAC < 18000 &
    nCount_RNA < 23000 &
    nCount_ATAC > 1000 &
    nCount_RNA > 1000 &
  #  nucleosome_signal < 2 &   ????? Should we know anything about that output?
    TSS.enrichment > 1
)
#pbmc

```

################################ PEAKS CALLING????

####THIS PART IS NOT WORKING!

# call peaks using MACS2
peaks <- CallPeaks(pbmc)
##### No troba res el peakcalling o que?

# remove peaks on nonstandard chromosomes and in genomic blacklist regions
peaks <- keepStandardChromosomes(peaks, pruning.mode = "coarse")
peaks <- subsetByOverlaps(x = peaks, ranges = blacklist_hg38_unified, invert = TRUE)

# quantify counts in each peak
macs2_counts <- FeatureMatrix(
  fragments=Fragments(pbmc),
  features = peaks,
  cells = colnames(pbmc)
)

# create a new assay using the MACS2 peak set and add it to the Seurat object
pbmc[["peaks"]] <- CreateChromatinAssay(
  counts = macs2_counts,
  fragments = fragpath,
  annotation = annotation
)




```{r}


############TREATING THE "COUNTS" AS PEAKS!
DefaultAssay(pbmc) <- "RNA"
pbmc <- SCTransform(pbmc)
pbmc <- RunPCA(pbmc)



DefaultAssay(pbmc) <- "ATAC"
pbmc <- FindTopFeatures(pbmc, min.cutoff = 5)
pbmc <- RunTFIDF(pbmc)
pbmc <- RunSVD(pbmc)


########ATACseqprova o que
DepthCor(pbmc)
pbmc <- RunUMAP(object = pbmc, reduction = 'lsi', dims = 2:30)
pbmc <- FindNeighbors(object = pbmc, reduction = 'lsi', dims = 2:30)
pbmc <- FindClusters(object = pbmc, verbose = FALSE, algorithm = 3, resolution = .05)
DimPlot(object = pbmc, label = TRUE, group.by = "cell_type") + NoLegend()



```



## Gene activities

```{r}



  #####GENE ACTIVITIES????
gene.activities <- GeneActivity(pbmc)



pbmc[['prova']] <- CreateAssayObject(counts = gene.activities)

#We normalize the data
pbmc <- NormalizeData(
  object = pbmc,
  assay = 'prova',
  normalization.method = 'LogNormalize',
  scale.factor = median(pbmc$nCount_prova)
)


pbmc <- NormalizeData(
  object = pbmc,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(pbmc$nCount_RNA)
)


#MALAT1, RPLP1, RPS27
#SLC8A1, UBE2E2, ZEB2
#We set default to RNA in order to plot for example
DefaultAssay(pbmc) <- 'RNA'

#######Mola que es pugui separar entre diferents layers???? o que.
#pbmc[['RNA']] <- split(pbmc[["RNA"]], f = pbmc$cell_type)

plot1 <- FeaturePlot(
  object = pbmc,
  features = c('CIITA', 'UBE2E2', 'ZEB2'),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3
)


#We set default to RNA in order to plot for example
DefaultAssay(pbmc) <- 'prova'

plot2 <- FeaturePlot(
  object = pbmc,
  features = c('CIITA', 'UBE2E2', 'ZEB2'),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3
)


plot1 | plot2
##Això ho podem comparar / predir o alguna vaina loca o que???'




```



```{r}

DefaultAssay(pbmc) <- "ATAC"

# first compute the GC content for each peak
pbmc <- RegionStats(pbmc, genome = BSgenome.Hsapiens.UCSC.hg38)



#PIK3AP1
#PHF21A



# link peaks to genes
pbmc <- LinkPeaks(
  object = pbmc,
  peak.assay = "ATAC",
  expression.assay = "SCT",
  genes.use = c("CIITA", "UBE2E2", "ZEB2")
)


#DimPlot(pbmc, group.by = "cell_type")


Idents(pbmc) <- pbmc@meta.data$cell_type







p1 <- CoveragePlot(
  object = pbmc,
  region = "CIITA",
  features = "CIITA",
  expression.assay = "SCT",
  #idents = idents.plot,
  extend.upstream = 500,
  extend.downstream = 1000
)

p2 <- CoveragePlot(
  object = pbmc,
  region = "UBE2E2",
  features = "UBE2E2",
  expression.assay = "SCT",
  #idents = idents.plot,
  extend.upstream = 800,
  extend.downstream = 500
)

p3 <- CoveragePlot(
  object = pbmc,
  region = "ZEB2",
  features = "ZEB2",
  expression.assay = "SCT",
  #idents = idents.plot,
  extend.upstream = 800,
  extend.downstream = 500
)

#patchwork::wrap_plots(p1, p2, p3, ncol = 1)
p1
p2
p3


```

```{r}
saveRDS(pbmc, file = "/home/roger/Github/OpenProblems/Signac/atacopen.rds")


pbmc <- readRDS(file = "/home/roger/Github/OpenProblems/Signac/atacopen.rds")
```



## MOTIF ANALYSIS

```{r}

library(Seurat)
library(Signac)
library(motifmatchr)
library(AnnotationDbi)
library(DirichletMultinomial)
library(JASPAR2020)
library(TFBSTools)
library(BSgenome.Hsapiens.UCSC.hg38)
library(patchwork)
library(BSgenome.Hsapiens.UCSC.hg19)

```
```{r}
```

Add list of motifs from JASPAR

```{r}

devtools::install_github("stuart-lab/signac", ref = "develop", auth_token="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")
install.packages("Signac")


library(Signac)
# Get a list of motif position frequency matrices from the JASPAR database
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(species=9606, all_versions=FALSE)
)


# opts = list(species = 9606) # 9606 is the species code for human
#list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
DefaultAssay(pbmc) <- 'ATAC'
# add motif information
pbmc <- AddMotifs(
  object = pbmc,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm = pfm
)


pbmc <- Footprint(
  object = pbmc,
  motif.name = c("MA0030.1", "MA0031.1"),
  genome = BSgenome.Hsapiens.UCSC.hg38,
  upstream = 10000,
  downstream=10000
)

InsertionBias(object=pbmc, genome=BSgenome.Hsapiens.UCSC.hg38, region = "chr1-1-249250621", verbose = TRUE)



```


```{r}

Idents(pbmc) <- pbmc@meta.data$cell_type

da_peaks <- FindMarkers(
  object = pbmc,
  ident.1 = 'T cells CD8+',
  ident.2 = 'T regulatory cells',
  only.pos = TRUE,
  test.use = 'LR',
  min.pct = 0.05,
  latent.vars = 'nCount_ATAC'
)

# get top differentially accessible peaks
top.da.peak <- rownames(da_peaks[da_peaks$p_val < 0.005, ])


```


OPTIONAL 
```{r}
open.peaks <- AccessiblePeaks(pbmc, idents = c("T cells CD4+", "B cells"))

meta.feature <- GetAssayData(pbmc, assay = "ATAC", slot = "meta.features")

peaks.matched <- MatchRegionStats(
  meta.feature = meta.feature[open.peaks, ],
  query.feature = meta.feature[top.da.peak, ],
  n = 50000
)

```




```{r}
# test enrichment
enriched.motifs <- FindMotifs(
  object = pbmc,
  features = top.da.peak
)
```

```{r}

library(ggseqlogo)
MotifPlot(
  object = pbmc,
  motifs = head(rownames(enriched.motifs))
)
```


```{r}



library(chromVAR)

pbmc <- RunChromVAR(
  object = pbmc,
  genome = BSgenome.Hsapiens.UCSC.hg38
)

DefaultAssay(pbmc) <- 'chromvar'

p1 <- DimPlot(pbmc, label = TRUE, pt.size = 0.1) + NoLegend()
p1
# look at the activity of Mef2c
p2 <- FeaturePlot(
  object = pbmc,
  features = "MA1513.1",
  min.cutoff = 'q10',
  max.cutoff = 'q90',
  pt.size = 0.1
)


#MA0497.1
p1 + p2
```



```{r}

library(presto)
differential.activity <- FindMarkers(
  object = pbmc,
  ident.1 = 'T cells CD8+',
  ident.2 = 'T regulatory cells',
  only.pos = TRUE,
  mean.fxn = rowMeans,
  fc.name = "avg_diff"
)

MotifPlot(
  object = pbmc,
  motifs = head(rownames(differential.activity)),
  assay = 'ATAC'
)


```



```{r}
DefaultAssay(pbmc) <- "ATAC"



pbmc <- Footprint(
  object = pbmc,
  motif.name = c("KLF15", "EGR1"),
  genome = BSgenome.Hsapiens.UCSC.hg38
)



# plot the footprint data for each group of cells
p2 <- PlotFootprint(pbmc, features = c("NRF1", "TCFL5", "HINFP"))

```





